MG7: Configurable and scalable 16S data analysis

### 2016-08-17 docs/mg7-paper/out/paper.tex
address @rtobes review in #88

Our 16S-DB7 Reference Database is a curated subset of sequences from the
NCBI nucleotide database \textbf{nt}. The sequences included were
selected by similarity with the bacterial and archaeal reference
sequences downloaded from the \textbf{RDP database}
\citep{cole2013ribosomal}. RDP unaligned sequences were used to capture
new 16S RNA sequences from \textbf{nt} using BLAST similarity search
strategies and then, performing additional curation steps to remove
sequences with poor taxonomic assignments to taxonomic nodes close to
the root of the taxonomy tree. All the nucleotide sequences included in
\textbf{nt} database has a taxonomic assignment provided by the
\textbf{Genbank} sequence submitter. NCBI provides a table (available at
ftp://ftp.ncbi.nlm.nih.gov/pub/taxonomy/) to do the mapping of any
Genbank Identifier (GI) to its Taxonomy Identifier (TaxID). Thus, we are
based on a crowdsourced submitter-maintained taxonomic annotation system
for reference sequences. It supposes a sustainable system able to face
the expected number of reference sequences that will populate the public
global nucleotide databases in the near future. Another advantageous
point is that we are based on NCBI taxonomy, the \emph{de-facto}
standard taxonomic classification for biomolecular data
\citep{cochrane20102010}. NCBI taxonomy is, undoubtedly, the most used
taxonomy all over the world and the most similar to the official
taxonomies of each specific field. This is a crucial point because all
the type-culture and tissue databanks follow this official taxonomical
classification and, in addition, all the knowledge accumulated during
last decades is referred to this taxonomy. In addition NCBI provides a
direct connection between taxonomical formal names and the physical
specimens that serve as exemplars for the species
\citep{federhen2014type}.

Certainly, if metagenomics results are easily integrated with the
theoretical and experimental knowledge of each specific area, the impact
of metagenomics will be higher than if it progresses as a disconnected
research branch. Considering that metagenomics data interoperability,
which is especially critical in clinical environments, requires a stable
taxonomy to be used as reference, we decided to rely on the most widely
used taxonomy: the NCBI taxonomy. In addition, the biggest global
sequence database GenBank follows this taxonomy to register the origin
of all their submitted sequences. Our 16S database building strategy
allows the substitution of the 16S database by any other subset of
\textbf{nt}, even by the complete \textbf{nt} database if it would be
needed, for example, for analyzing shotgun metagenomics data. This
possibility of changing the reference database provides flexibility to
the system enabling it for easy updating and project-driven
personalization.

>>>>

The starting point for our 16S reference database is RNA Central
\citep{rnacentral2014rnacentral},
\href{http://blog.rnacentral.org/2016/03/rnacentral-release-5.html}{version
5}. RNACentral was chosen being the most up to date, comprehensive RNA
sequence repository, including among others all RNA data from Silva,
GreenGenes, RDP, ENA, and RefSeq. First we take those sequences which

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  are annotated as being of \texttt{rRNA} type\footnote{We are aware of
    the existence of a gene annotation corresponding to 16S in
    RNACentra, that we are \textbf{not using} due to a significant
    amount of 16S sequences lacking it}
\item
  their length is at least \texttt{1300}\footnote{Note that 16S
    sequences are sometimes part of an entry corresponding to whole
    small subunits; that's why we do not set a maximum length threshold}
\item
  have at least one taxonomic assignment to a descendant of
  \emph{Bacteria} or \emph{Archaea}
\item
  their lineage does not contain a set of taxa deemed
  uninformative\footnote{for example: ``unclassified Bacteria
    (miscellaneous)'' or ``unclassified''. Assigning a read to an
    ``unclassified'' taxon defeats the first and foremost goal of
    taxonomic profiling: \emph{classification}.}
\end{enumerate}

After this first step, we drop redundant assignments: if sequences
\(S_1 \subseteq S_2\) share an assignment \(T\), it gets dropped from
\(S_1\); sequences which as a result of this process end up having no
assignments are removed.

Once we have a non-redundant set of ribosomal RNA sequences (containing
all those corresponding to 16S), we can apply our procedure for checking
internal consistency of taxonomic assignments. We run MG7 using this set
of sequences as query, with reference all but the one we are trying to
assign. Under the (reasonable) assumption of 16S sequence similarity
being correlated with the taxonomy tree topology, the resulting MG7
taxonomic assignment should either

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  be a taxon \emph{close} to the original assignment, if there are
  sequences similar enough in the reference database; what we consider
  as close is: the parent of the MG7 assignment should be contained in
  the lineage of the original assignment.
\item
  be empty, due to this sequence having no similar sequences in the
  reference database, thus making impossible to do any taxonomic
  assignment.
\end{enumerate}

In any other case, under our assumptions, this assignment should be
discarded.

All these steps are automated, and integrated with every release of our
reference database. The whole process can be repeated and adapted to
other subsets of RNACentral; we do this for an 18S reference database.
